---
title: "Phenotype diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PhenotypeDiagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")

knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
eval = NOT_CRAN
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(CDMConnector)
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!eunomiaIsAvailable()) downloadEunomiaData(datasetName = "synpuf-1k", cdmVersion = "5.3")
```

## Introduction: Run PhenotypeDiagnostics

In this vignette, we are going to present how to run `PhenotypeDiagnostics()`. We are going to use the following packages and mock data:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(CohortConstructor)
library(PhenotypeR)
library(dplyr)

con <- DBI::dbConnect(duckdb::duckdb(), 
                      CDMConnector::eunomiaDir("synpuf-1k", "5.3"))
cdm <- CDMConnector::cdmFromCon(con = con, 
                                cdmName = "Eunomia Synpuf",
                                cdmSchema   = "main",
                                writeSchema = "main", 
                                achillesSchema = "main")
cdm
```

Note that we have included [achilles tables](https://github.com/OHDSI/Achilles) in our cdm reference, which will be used to speed up some of the analyses.

## Create a cohort

First, we are going to use the package [CohortConstructor](https://ohdsi.github.io/CohortConstructor/) to generate three cohorts of *warfarin*, *acetaminophen* and *morphine* users.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create a codelist
codes <- list("warfarin" = c(1310149, 40163554),
              "acetaminophen" = c(1125315, 1127078, 1127433, 40229134, 40231925, 40162522, 19133768),
              "morphine" = c(1110410, 35605858, 40169988))

# Instantiate cohorts with CohortConstructor
cdm$my_cohort <- conceptCohort(cdm = cdm,
                               conceptSet = codes, 
                               exit = "event_end_date",
                               overlap = "merge",
                               name = "my_cohort")
```
## Run PhenotypeDiagnostics

Now we will proceed to run `phenotypeDiagnotics()`. This function will run the following analyses:

-   **Database diagnostics**: This includes information about the size of the data, the time period covered, the number of people in the data, and other meta-data of the CDM object. See [Database diagnostics vignette](https://ohdsi.github.io/PhenotypeR/articles/a03_DatabaseDiagnostics.html) for more details. 
-   **Codelist diagnostics**: This includes information on the concepts included in our cohorts' codelist. See [Codelist diagnostics vignette](https://ohdsi.github.io/PhenotypeR/articles/a04_CodelistDiagnostics.html) for further details.
-   **Cohort diagnostics**: This summarises the characteristics of our cohorts, as well as comparing them to age and sex matched controls from the database.. See [Cohort diagnostics vignette](https://ohdsi.github.io/PhenotypeR/articles/a05_CohortDiagnostics.html) for further details.
-   **Population diagnostics**: Calculates the frequency of our study cohorts in the database in terms of their incidence rates and prevalence. See [Population diagnostics vignette](https://ohdsi.github.io/PhenotypeR/articles/a07_PopulationDiagnostics.html) for further details.

### Default arguments
If we do not provide any specification, the default values of the functions will be used. That means, the following
two scripts provide the same summarised result:

```{r, eval=FALSE}
result1 <- phenotypeDiagnostics(cdm$warfarin,
                                databaseDiagnostics = list(),
                                codelistDiagnostics = list(),
                                cohortDiagnostics = list(),
                                populationDiagnostics = list())

result2 <- phenotypeDiagnostics(cdm$warfarin,
                                databaseDiagnostics = list(
                                  "diagnostics" = c("snapshot", "personTableSummary",
                                  "observationPeriodsSummary", "clinicalRecordsSummary")
                                ),
                                codelistDiagnostics = list(
                                  "diagnostics" = c("achillesCodeUse", "orphanCodeUse",
                                                    "cohortCodeUse", "drugDiagnostics",
                                                    "measurementDiagnostics"),
                                  "measurementDiagnosticsSample" = 20000,
                                  "drugDiagnosticsSample" = 20000
                                ),
                                cohortDiagnostics = list(
                                  "diagnostics" = c("cohortCount", "cohortCharacteristics",
                                                    "largeScaleCharacteristics",
                                                    "compareCohorts"),
                                  "cohortSample" = 20000,
                                  "matchedSample" = 1000
                                ),
                                populationDiagnostics = list(
                                  "diagnostics" = c("incidence", "periodPrevalence"),
                                  "populationSample" = 100000,
                                  "populationDateRange" = as.Date(c(NA,NA))
                                ))
```
Notice that the default values do not include running 'cohortSurvival' analysis in CohortDiagnostics. 
If we want to add it, we just need to add the analysis in the CohortDiagnostics "diagnostics" argument:

```{r, eval=FALSE}
result3 <- phenotypeDiagnostics(cdm$warfarin,
                               databaseDiagnostics = list(
                                 "diagnostics" = c("snapshot", "personTableSummary",
                                                   "observationPeriodsSummary", "clinicalRecordsSummary")
                               ),
                               codelistDiagnostics = list(
                                 "diagnostics" = c("achillesCodeUse", "orphanCodeUse",
                                                   "cohortCodeUse", "drugDiagnostics",
                                                   "measurementDiagnostics"),
                                 "measurementDiagnosticsSample" = 20000,
                                 "drugDiagnosticsSample" = 20000
                               ),
                               cohortDiagnostics = list(
                                 "diagnostics" = c("cohortCount", "cohortCharacteristics",
                                                   "largeScaleCharacteristics",
                                                   "compareCohorts", "cohortSurvival"),
                                 "cohortSample" = 20000,
                                 "matchedSample" = 1000
                               ),
                               populationDiagnostics = list(
                                 "diagnostics" = c("incidence", "periodPrevalence"),
                                 "populationSample" = 100000,
                                 "populationDateRange" = as.Date(c(NA,NA))
                               ))
```

### Skipping diagnostics or sub-analysis
To skip a diagnostics, we just need to specify that one as NULL. For instance, in
the following example, **databaseDiagnostics** will not be run:

```{r, eval=FALSE}
result4 <- phenotypeDiagnostics(
  cohort = cdm$my_cohort, 
  databaseDiagnostics = NULL)
```

Or, if for instance, we would like to run **databaseDiagnostics** but only run the
*snapshot* analysis, we can specify it like this:
```{r, eval=FALSE}
result4 <- phenotypeDiagnostics(
  cohort = cdm$my_cohort, 
  databaseDiagnostics = list(
    "diagnostics" = "snapshot"
  ))
```

## Save the results
To save the results, we can use [exportSummarisedResult](https://darwin-eu.github.io/omopgenerics/reference/exportSummarisedResult.html) function from [omopgenerics](https://darwin-eu.github.io/omopgenerics/index.html) R Package:
```{r, eval=FALSE}
exportSummarisedResult(result3, path = here::here(), minCellCount = 5)
```

## Visualisation of the results
Once we get our **Phenotype diagnostics** result, we can use `shinyDiagnostics` to easily create a shiny app and visualise our results:

```{r, eval=FALSE}
shinyDiagnostics(result,
                 directory = tempdir(),
                 minCellCount = 5, 
                 open = TRUE)
```
Notice that we have specified the minimum number of counts (`minCellCount`) for suppression to be shown in the shiny app, and also that we want the shiny to be launched in a new R session (`open`). You can see the shiny app generated for this example in [here](https://dpa-pde-oxford.shinyapps.io/PhenotypeRShiny/).See [Shiny diagnostics vignette](https://ohdsi.github.io/PhenotypeR/articles/a02_ShinyDiagnostics.html) for a full explanation of the shiny app.

