---
title: "Population diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PopulationDiagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")

knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
eval = NOT_CRAN
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7
)

library(omock)
if (Sys.getenv("OMOP_DATA_FOLDER") == "") Sys.setenv("OMOP_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("OMOP_DATA_FOLDER"))) dir.create(Sys.getenv("OMOP_DATA_FOLDER"))
```

## Introduction
In this example we're going to just create a cohort of individuals with an ankle sprain using the Eunomia synthetic data. 

```{r, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(CodelistGenerator)
library(PatientProfiles)
library(IncidencePrevalence)
library(PhenotypeR)

cdm <- omock::mockCdmFromDataset(datasetName = "synpuf-1k_5.3", source = "duckdb")

cdm$injuries <- conceptCohort(cdm = cdm,
  conceptSet = list(
    "ankle_sprain" = 81151
  ),
  name = "injuries")
```

We can get the incidence and prevalence of our study cohort using `populationDiagnostics()`:
```{r}
pop_diag <- populationDiagnostics(cdm$injuries)
```

This function builds on [IncidencePrevalence](https://darwin-eu.github.io/IncidencePrevalence/index.html) R package to perform the following analyses:

-   **Incidence:** It estimates the incidence of our cohort using [estimateIncidence()](https://darwin-eu.github.io/IncidencePrevalence/reference/estimateIncidence.html).
-   **Period Prevalence:** It estimates the period prevalence of our cohort on a year basis using [estimatePeriodPrevalence()](https://darwin-eu.github.io/IncidencePrevalence/reference/estimatePeriodPrevalence.html).

We can deactivate any of the analysis turning the following arguments into FALSE:

```{r}
pop_diag2 <- populationDiagnostics(cdm$injuries, 
                                  incidence = TRUE, 
                                  periodPrevalence = TRUE)
```

All analyses are performed for:

-   Overall and stratified by age groups: 0 to 17, 18 to 64, 65 to 150. Age groups cannot be modified.
-   Overall and stratified by sex (Female, Male).
-   Restricting the denominator population to those with 0 and 365 of days of prior observation. Prior observation values cannot be modified.

As calculating the incidence and prevalence of our cohorts can be computationally expensive, specially when working with large databases, we can subset our diagnostics to a subset of people using `populationSample`, and restrict the study period to specific dates using `populationDateRange`:
```{r}
pop_diag3 <- populationDiagnostics(cdm$injuries, 
                                   populationSample = 1000, 
                                   populationDateRange = as.Date(c("1960-01-01", NA)))
```


## Visualising the results
We can use  [IncidencePrevalence](https://darwin-eu.github.io/IncidencePrevalence/index.html) package to visualise the results obtained.

### Incidence
```{r}
tableIncidence(pop_diag,     
               groupColumn = c("cdm_name", "outcome_cohort_name"),
               hide = "denominator_cohort_name",
               settingsColumn = c("denominator_age_group",
                                  "denominator_sex",
                                  "denominator_days_prior_observation",
                                  "outcome_cohort_name"))
```

```{r}
results <- pop_diag |> 
  omopgenerics::filterSettings(result_type == "incidence") |>
  visOmopResults::filterAdditional(analysis_interval == "years")
plotIncidence(results,
              colour = "denominator_age_group",
              facet = c("denominator_sex", "denominator_days_prior_observation"))
```


### Prevalence
```{r}
tablePrevalence(pop_diag,     
                groupColumn = c("cdm_name", "outcome_cohort_name"),
                hide = "denominator_cohort_name",
                settingsColumn = c("denominator_age_group",
                                   "denominator_sex",
                                   "denominator_days_prior_observation",
                                   "outcome_cohort_name"))
```

```{r}
results <- pop_diag |> 
  omopgenerics::filterSettings(result_type == "prevalence") |>
  visOmopResults::filterAdditional(analysis_interval == "years")
plotPrevalence(results,
               colour = "denominator_age_group",
               facet = c("denominator_sex", "denominator_days_prior_observation"))
```
